DynaProg for Scala

### 2013-01-17 docs/report/benchmarks.tex
Nits. Bug in the backtrack of two track grammars.

\documentclass[11pt]{article}
\input{inc/style.sty}
\usepackage{multicol}
\usepackage{etoolbox}

\title{DynaProg for Scala}
\subtitle{A Scala DSL for Dynamic Programming on CPU and GPU}
\begin{document}
\maketitle
\input{introduction}
\input{problems}
\input{design}
\input{implementation}

\section{Usage} \label{usage}

@@

\subsection{Program examples}
\subsubsection{Matrix multiplication} \label{ex_matmult}
\begin{lstlisting}[language=Scala,caption={Matrix Multiplication}]
import v4._

@@

\subsubsection{RNA folding} \label{ex_rnafold}
\begin{lstlisting}[language=Scala,caption={RNA folding}]
import v4._

trait RNAFoldSig extends RNASignature {
  def hairpin(ij:(Int,Int)):Answer
  def stack(i:Int,s:Answer,j:Int):Answer
  def iloop(ik:(Int,Int),s:Answer,lj:(Int,Int)):Answer
  def mloop(i:Int,s:Answer,j:Int):Answer
  def left(l:Answer,r:Int):Answer
  def right(l:Int,r:Answer):Answer
  def join(l:Answer,r:Answer):Answer
}

trait RNAFoldAlgebra extends RNAFoldSig {
  type Answer = Int
  import librna.LibRNA._ // indexing convention: first base,last base
  private def Eh(i:Int,j:Int) = hl_energy(i,j) // hairpin loop
  private def Ei(i:Int,k:Int,l:Int,j:Int) = il_energy(i,k,l,j) // internal loop
  private def Es(i:Int,j:Int) = sr_energy(i,j) // stacked base pairs
  def hairpin(ij:(Int,Int)) = Eh(ij._1,ij._2-1)
  def stack(i:Int,s:Int,j:Int) = Es(i,j) + s
  def iloop(ik:(Int,Int),s:Int,lj:(Int,Int)) = Ei(ik._1,ik._2,lj._1-1,lj._2-1) + s
  def mloop(i:Int,s:Int,j:Int) = s
  def left(l:Int,r:Int) = l
  def right(l:Int,r:Int) = r
  def join(l:Int,r:Int) = l+r
  override val h = min[Answer] _
}

trait RNAFoldPrettyPrint extends RNAFoldSig {
  type Answer = String
  private def dots(n:Int,c:Char='.') = (0 until n).map{_=>c}.mkString
  def hairpin(ij:(Int,Int)) = "("+dots(ij._2-ij._1-2)+")"
  def stack(i:Int,s:String,j:Int) = "("+s+")"
  def iloop(ik:(Int,Int),s:String,lj:(Int,Int)) = "("+dots(ik._2-1-ik._1)+s+dots(lj._2-1-lj._1)+")"
  def mloop(i:Int,s:String,j:Int) = "("+s+")"
  def left(l:String,r:Int) = l+"."
  def right(l:Int,r:String) = "."+r
  def join(l:String,r:String) = l+r
}

trait RNAFoldGrammar extends ADPParsers with RNAFoldSig {
  lazy val Qp:Tabulate = tabulate("Qp",(
    seq(3,maxN)        ^^ hairpin
  | eli   ~ Qp ~ eli   ^^ stack
  | seq() ~ Qp ~ seq() ^^ iloop
  | eli   ~ QM ~ eli   ^^ mloop
  ) filter basepairing aggregate h)

  override val axiom = Q
}

object RNAFold extends App {
  object fold extends RNAFoldGrammar with RNAFoldAlgebra
  object pretty extends RNAFoldGrammar with RNAFoldPrettyPrint

  val seq="aaaaaagggaaaagaacaaaggagacucuucuccuuu"

  import librna.LibRNA._
  setSequence(seq)
  val (score,bt) = fold.backtrack(seq.toArray).head
  val res = pretty.build(seq.toArray,bt)
  clear
  println("Folding : "+res+" (%5.2f)".format(score/100.0));
}
\end{lstlisting}

@@

\section{Conclusion}
{\color{red} XXX}

\begin{verbatim}

recap of the report

recap of the results

big picture: with actual compiler technology we 
Embedded DSL vs external DSL
Larger picture how to go further

\end{verbatim}

% ------------------------------------------------------------------------------------------------
\newpage
\section*{Planning}
\subsubsection*{Todo @TCK}\ul
\item Test/proof parsers are correct -- make sure implementation is correct
\item Automate test to compare against implementation
\item Benchmarks -- use CUDA profiler(?)
\item Write report
\item Port LibRNA for CUDA?
\ule

\subsubsection*{Todo @Manohar}\ul
\item Integrate LMS code generation into v4.
\item Fix Zuker coefficients
\ule

Legacy roadmap deadlines: \\
\begin{tabular}{ll}
Nov 16 & Rules normalization and automatic backtracking \\
	& GenScala on LMS + GenCuda + LMS CudaCompiler \\
Nov 23 & Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization \\
Nov 30 &--- gap due to LMS missing knowledge --- \\
Dec 7 & Benchmarking, grammar analysis \\
Dec 14 & First thoughts for larger than device memory \\
Dec 21 & Writing report \\
Jan 4 & --- holiday --- \\
Jan 18 & Writing report: implementation description and plan for future work
\end{tabular}

\subsubsection*{Journal}
\begin{tabular}{ll}
Sep. 17	& Getting started with the project, reading related work on hash maps. \\
Sep. 24	& Parallel hashing paper solve the problem for 32bit key/value pair. Stripped CudPP. \\
		& Devised (but not implemented) extension beyond 64bit using memory areas locks. \\
---		& Change of project suggested by Vojin (supervisor): joint work with Manohar on DP \\
Oct. 01	& Problems specifications: serial/non-serial, started CUDA implementation with blocks. \\
Oct. 08	& CudAlign solves serial monadic, might adapt it (but complicated / ad-hoc / in progress). \\
		& Focus on non-serial problems that fit in GPU memory. \\
Oct. 15	& First working implementation for non-serial problems (rectangle, triangle, parallelogram) \\
Oct. 22	& No workaround for timeout, fixed by multiple kernels. Implemented backtrack on GPU. \\
Oct. 29	& Scala/C compiler engine, to use Scala/CUDA, C code must be provided. ADP (Manohar) \\
Nov. 05	& Multiple fixes and rework of ADP parsers to aim at generating C-like code. \\
Nov. 12	& Rework ADP parsers: cyclic, two-track grammars and automatic backtracking discussion. \\
Nov. 19	& Explorations LMS and Macros for C code, implementation is quite different, hence ad-hoc. \\
Nov. 26	& Full backtracking stack: apply / unapply / reapply, refactoring of the classes. \\
Dec. 03	& JNI for LibRNA to get coefficients for Zuker, errors in algebra (quite hairy code). \\
Dec. 10	& Rework concatenation operators, yield analysis, code generation. \\
Dec. 17	& Detupling, generic backtrack (vs. ad-hoc), nested aggregates, empty results support. \\
Dec. 24	& (sick 4 days), code generation: full JNI conversion support, rework CodeCompiler. \\
Dec. 31	& Writing report (design, implementation), fixing various issues, complexity analysis. \\
Jan. 7	& Writing report, attempt to fix Zuker, fix SBT, strip LibRNA. LMS C codegen (Manohar) \\
Jan. 14	& {\color{red} XXX} \\
\end{tabular}

%Some GPU algorithms: http://hgpu.org/?cat=11
%Translation into C++: http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/
%CUDPP libraries (but awfully big resulting binary): http://code.google.com/p/cudpp/
%13 dwarfs: http://developer.amd.com/afds/assets/presentations/2155_final.pdf
%http://tutorials.jenkov.com/java-reflection/fields.html
%http://lampwww.epfl.ch/~michelou/scala/scala-reflection.html
%
%Hint: use TypeClass to put a predicate on types
%  def fun[T: CanTranslateToC](...)
%  def fun[T](implicit ev:CanTranslateToC[T])
%  class CanTranslateToC[T] { def translate:String }
%  implicit def canTranslateInt = new CanTranslateToC[Int] = { def translate = "Int" }

% ------------------------------------------------------------------------------------------------
\newpage
\patchcmd{\thebibliography}{\section*{\refname}}{\begin{multicols}{2}[\section*{\refname}]}{}{}
\patchcmd{\endthebibliography}{\endlist}{\endlist\end{multicols}}{}{}
\bibliographystyle{plain} % acm
{\small \bibliography{bibliography.bib}}
\end{document}

>>>>

\newpage
\section{Usage} \label{usage}
\subsection{Program examples}
In this section, we explain how to use DynaProg, as a parser library. Use an example based approach, we focus on three additional examples: Smith-Waterman/Needleman-Wunsch to present two-tracks grammars and multiple algebras, RNAfold\cite{gpu_rnafold} to describe RNA library usage and reconstruction from backtrack trace, and finally we extend Matrix chain multiplication (\S\ref{adp_practice}) with CUDA code generation.

\subsubsection{Smith-Waterman and Needleman-Wunsch} \label{ex_swat}
We first define a signature that can fit both algebrae, then we specify for each algebra the functions. In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general, see \S\ref{ex_rnafold}).

We also want to emphasize how bad this example is, with respect to the result construction, as every intermediate result would contain its whole backtrack, hereby duplicating the information multiple times. To overcome this issue {\color{red} XXX}
\begin{lstlisting}[language=Scala,captionpos=none]
trait SeqAlignSignature extends Signature {
  type Alphabet = Char
  type Answer = (Int,String,String) // score, seq1, seq2
  override val h = maxBy((x:Answer)=>x._1)

  def gap(current:Int, g:(Int,Int)):Int
  def pair(a:Alphabet,b:Alphabet):Int
}

trait SmithWatermanAlgebra extends SeqAlignSignature {
  private val open = -3
  private val extend = -1
  def gap(score:Int, g:(Int,Int)) = { val size=g._2-g._1; Math.max(0, score + ( open + (size-1) * extend )) }
  def pair(a:Char,b:Char) = if (a==b) 10 else -3
}

trait NeedlemanWunschAlgebra extends SeqAlignSignature {
  private val open = -15
  private val extend = -1
  def gap(score:Int, g:(Int,Int)) = { val size=g._2-g._1; score + ( open + (size-1) * extend ) }
  def pair(a:Char,b:Char) = if (a==b) 4 else -3
}
\end{lstlisting}

\begin{lstlisting}[language=Scala,captionpos=none]
trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
  private def prettyGap(sw:Subword,in:Function1[Int,Char]):(String,String) = {
    val g=(sw._1 until sw._2).toList; (g.map{x=>in(x)}.mkString(""),g.map{x=>"-"}.mkString(""))
  }

  val alignment:Tabulate = tabulate("M",(
    empty                         ^^ { _ => (0,".",".") }
  | seq1() -~ alignment           ^^ { case (g,(score,s1,s2)) => val (g1,g2)=prettyGap(g,in1); (gap(score,g),s1+g1,s2+g2) }
  |           alignment ~- seq2() ^^ { case ((score,s1,s2),g) => val (g1,g2)=prettyGap(g,in2); (gap(score,g),s1+g2,s2+g1) }
  | el1    -~ alignment ~- el2    ^^ { case (c1,((score,s1,s2),c2)) => (score+pair(c1,c2),s1+c1, s2+c2) }
  ) aggregate h)

  val axiom = alignment

  def align(s1:String,s2:String) = parse(s1.toArray,s2.toArray).head
}

object SeqAlign extends App {
  object SWat extends SeqAlignGrammar with SmithWatermanAlgebra   // Smith-Waterman
  object NWun extends SeqAlignGrammar with NeedlemanWunschAlgebra // Needleman-Wunsch
  val seq1 = "CGATTACA"
  val seq2 = "CCCATTAGAG"

  val (swScore,sw1,sw2) = SWat.align(seq1,seq2)
  println("Smith-Waterman alignment\n- Score: "+swScore+"\n- Seq1: "+sw1+"\n- Seq2: "+sw2+"\n")
  val (nwScore,nw1,nw2) = NWun.align(seq1,seq2)
  println("Needleman-Wunsch alignment\n- Score: "+nwScore+"\n- Seq1: "+nw1+"\n- Seq2: "+nw2+"\n")
}
\end{lstlisting}

\newpage
\subsubsection{RNA folding} \label{ex_rnafold}
We define a signature with two evaluation algebras: {\tt RNAFoldAlgebra} actually computes the folding whereas {\tt RNAFoldPrettyPrint} describes the folding in a string. The energy functions are provided by an external library ({\tt LibRNA}). This library encodes substring as (first character, last character) whereas our framework encodes them as (first character, first character + length), which explains the off-by-one corrections. {\tt energies} variable is set to false in {\tt RNAFoldPrettyPrint} because this algebra does not involve the LibRNA energies functions (that require encoding the input RNA sequence in a special format; this option is enabled by default in the RNASignature trait).

\begin{lstlisting}[language=Scala,captionpos=none]
trait RNAFoldSig extends RNASignature {
  def hairpin(ij:(Int,Int)):Answer
  def stack(i:Int,s:Answer,j:Int):Answer
  def iloop(ik:(Int,Int),s:Answer,lj:(Int,Int)):Answer
  def mloop(i:Int,s:Answer,j:Int):Answer
  def left(l:Answer,r:Int):Answer
  def right(l:Int,r:Answer):Answer
  def join(l:Answer,r:Answer):Answer
}

trait RNAFoldAlgebra extends RNAFoldSig {
  type Answer = Int
  import librna.LibRNA._ // indexing convention: first base,last base
  def hairpin(ij:(Int,Int)) = hl_energy(ij._1,ij._2-1) // Eh
  def stack(i:Int,s:Int,j:Int) = sr_energy(i,j) + s // Es
  def iloop(ik:(Int,Int),s:Int,lj:(Int,Int)) =
      il_energy(ik._1,ik._2,lj._1-1,lj._2-1) + s // Ei
  def mloop(i:Int,s:Int,j:Int) = s
  def left(l:Int,r:Int) = l
  def right(l:Int,r:Int) = r
  def join(l:Int,r:Int) = l+r
  override val h = min[Answer] _
}

trait RNAFoldPrettyPrint extends RNAFoldSig {
  type Answer = String
  override val energies=false
  private def dots(n:Int,c:Char='.') = (0 until n).map{_=>c}.mkString
  def hairpin(ij:(Int,Int)) = "("+dots(ij._2-ij._1-2)+")"
  def stack(i:Int,s:String,j:Int) = "("+s+")"
  def iloop(ik:(Int,Int),s:String,lj:(Int,Int)) =
      "("+dots(ik._2-1-ik._1)+s+dots(lj._2-1-lj._1)+")"
  def mloop(i:Int,s:String,j:Int) = "("+s+")"
  def left(l:String,r:Int) = l+"."
  def right(l:Int,r:String) = "."+r
  def join(l:String,r:String) = l+r
}
\end{lstlisting}

\newpage
We can then define the associated grammar
\begin{lstlisting}[language=Scala,captionpos=none]
trait RNAFoldGrammar extends ADPParsers with RNAFoldSig {
  lazy val Qp:Tabulate = tabulate("Qp",(
    seq(3,maxN)        ^^ hairpin
  | eli   ~ Qp ~ eli   ^^ stack
  | seq() ~ Qp ~ seq() ^^ iloop
  | eli   ~ QM ~ eli   ^^ mloop
  ) filter basepairing aggregate h)

  override val axiom = Q
}
\end{lstlisting}

In the application, we create two objects, each combining the grammar with a particular algebra. We can optionally specify a coefficient parameter file with {\tt setParams(file:String)}, otherwise the \textit{Turner2004} coefficients are used. The library is automatically loaded and fed with the sequence to produce correct energy coefficients. We request both the score and the backtrack trace (in {\tt bt}) so that we can reconstruct the folding using the pretty printing grammar.
\begin{lstlisting}[language=Scala,captionpos=none]
object RNAFold extends App {
  object fold extends RNAFoldGrammar with RNAFoldAlgebra
  object pretty extends RNAFoldGrammar with RNAFoldPrettyPrint

  val seq="aaaaaagggaaaagaacaaaggagacucuucuccuuuuucaaaggaagagg"

  val (score,bt) = fold.backtrack(seq.toArray).head
  val res = pretty.build(seq.toArray,bt)
  println("Folding : "+res+" (%5.2f)".format(score/100.0));
}
\end{lstlisting}

\newpage
\subsubsection{Matrix multiplication with CUDA code generation} \label{ex_matmult_cuda_lms}
Leveraging the existing definitions of the signature and grammar
\begin{lstlisting}[language=Scala,captionpos=none]
trait MatrixSig extends Signature {
  type Alphabet = (Int,Int) // Matrix(rows, columns)
  val single:Alphabet=>Answer
  val mult:(Answer,Answer)=>Answer
}

trait MatrixGrammar extends ADPParsers with MatrixSig {
  val axiom:Tabulate = tabulate("M",
    (el ^^ single | axiom ~ axiom ^^ mult) aggregate h)
}
\end{lstlisting}
We need describe the algebra functions in the LMS syntax ({\tt RepWorld}) that we can later compile to use as regular functions, augmented with C code description (necessary for code generation). Finally, we need to mix the {\tt CodeGen} trait to enable code generation and provide the manifest for input and ouput types ({\tt Alphabet} and {\tt Answer}).

\begin{lstlisting}[language=Scala,captionpos=none]
trait RepWorld extends NumericOps with TupleOps {
  type Alphabet = (Int, Int)
  type Answer = (Int, Int, Int)

  def hf(a: Rep[Answer]) :Rep[Int] = a._2
  def repSingle(a: Rep[Alphabet]): Rep[Answer] = (a._1, unit(0), a._2)
  def repMult(l: Rep[Answer], r: Rep[Answer]): Rep[Answer] =
    (l._1, l._2 + r._2 + l._1 * l._3 * r._3, r._3)
}

object MatrixMultLMS extends MatrixSig with MatrixGrammar
    with CodeGen with App {
  override val tps=(manifest[Alphabet],manifest[Answer])
  override val benchmark = true // display timing measurements

  // Algebra is defined immediately in the concrete program
  type Answer = (Int, Int, Int)
  val concreteProg = new RepWorld with RepPackage
  override val h = minBy(concreteProg.gen(concreteProg.hf))
  val single = concreteProg.gen(concreteProg.repSingle)
  val mult = concreteProg.gen2(concreteProg.repMult)

  val input = List((1,2),(2,20),(20,2),(2,4),(4,2),(2,1),(1,7),(7,3)).toArray
  println(parse(input).head) // -> 1x3 matrix, 122 multiplications
}
\end{lstlisting}

\newpage
----------------------------------------

 and focus on three different problem: our matrix multiplication example,

@@

\subsubsection{Matrix multiplication} \label{ex_matmult}
\begin{lstlisting}[language=Scala,caption={Matrix Multiplication}]
import v4._

@@

@@

### 2013-01-17 docs/report/introduction.tex
Nits. Bug in the backtrack of two track grammars.

% -----------------------------
\subsubsection{ADP in practice}
ADP is a formalization of parsers that introduces a distinction between the \textbf{parsing grammar} (recognition phase) and an associated \textbf{algebra} (evaluation phase). Such separation makes it possible to define multiple algebra for the same grammar. This has two main applications:\ol
\item Experiment variants with the same grammar:  for example, Needleman-Wunsch and Smith-Waterman share the same grammar but have a different evaluation algebra
\item Use an evaluation and execution algebra: a dynamic programming problem is solved in two steps: computing one optimal solution and applying it over actual data. For example in matrix chain multiplication, the first step solves the underlying dynamic program by evaluating the number of necessary multiplications, the second step \textit{effectively} multiplies matrices according to the order previously defined.
\ole

@@

\begin{lstlisting}[language=Scala, caption=Matrix chain mulitiplication DSL implementation]
trait MatrixSig extends Signature {
  type Alphabet = (Int,Int) // Matrix(rows, columns)
  val single:Alphabet=>Answer
  val mult:(Answer,Answer)=>Answer
}

>>>>

% -----------------------------
\subsubsection{ADP in practice} \label{adp_practice}
ADP is a formalization of parsers that introduces a distinction between the \textbf{parsing grammar} (recognition phase) and an associated \textbf{algebra} (evaluation phase). Such separation makes it possible to define multiple algebra for the same grammar. This has two main applications:\ol
\item Experiment variants with the same grammar:  for example, Needleman-Wunsch and Smith-Waterman share the same grammar but have a different evaluation algebra
\item Use an evaluation and execution algebra: a dynamic programming problem is solved in two steps: computing one optimal solution and applying it over actual data. For example in matrix chain multiplication, the first step solves the underlying dynamic program by evaluating the number of necessary multiplications, the second step \textit{effectively} multiplies matrices according to the order previously defined.
\ole

@@

\begin{lstlisting}[language=Scala,caption={Matrix chain mulitiplication DSL implementation}]
trait MatrixSig extends Signature {
  type Alphabet = (Int,Int) // Matrix(rows, columns)
  val single:Alphabet=>Answer
  val mult:(Answer,Answer)=>Answer
}
